\section{Mwana mubehi}\footnote{written in the orthography used by the informant}
Khale yaliho mwana muyayi lira lie Mbalaka yayilanga kuka wewe tsing’\-ombe.  Litukhu lindi bandu bahulila na lanjilitsanga ari bandu bitsi khuhuba tsing’ombe.  Bandu barula mutsinzu tsiabo khutsia khwilukhanyinya bihubi.  Bandu bene yabo bitsa nende marimu, tsindoboshi nende tsimpanga khukwana bihubi.  Shibanyola bihubi tawe.  Mbalaka yaranga khusakha na abola ndi yenyie khulola shia bandu bakhola.  Inyanga yindi khandi yalanjilitsa bandu bahubanga tsing’ombe.  Bandu khandi bitsa ni banyola Mbalaka anini mumusala asakhanga.  Lwa khabaka ya lanjilitsa bandu shibayitsa tawe shichila bapara ndinagwa ababehanga khuli matukhu kandi.  Mbalaka yalanjilitsa mpaka nachioka.  Mundu shiyayitsa tawe.  Ikhabi indamanu, litikhu yili bihubi biali ni bitsi nibihuba tsingo’mbe tsia kuka wewe.

\section{The boy who shouted}
A long time ago, there was a young child named Mbalaka, who was grazing cattle for his grandfather.  One day, people heard him shouting, that people had come to steal the cattle.  The people came from their huts to go chase away the thieves.  Those same people cam with spears and whips and machetes to fight the thieves.  They didn't find the thieves.  Mbalaka started to laugh, and told them that he had wanted to see what the people would do.  Another day, again he shouted that people are stealing the cattle.  The people again came, and found that Mbalaka had climbed up into a tree and was laughing.  The third time he shouted, the people did not come because they thought that he was lying to them, like the other days.  Mbalaka shouted until he became tired.  Not a person came.  Bad luck, for this day thieves had come and stole his grandfather's cattle.

\section{Analyzation}

\noindent Line 1: Original text\\
Line 2: Transcription with morpheme breaks\footnote{due to sound changes when morphemes combine, the transcription is not 100\% phonetic here}\\
Line 3: Morphemic analysis


\glll {Khale} {yaliho} {mwana} {muyayi} {lira} {lie} {Mbalaka} {yayilanga} {kuka} {wewe} {tsing’ombe.} {Litukhu} {lindi} {bandu} {bahulila} {na lanjilitsanga} {ari} {bandu} {bitsi} {khuhuba} {tsing’ombe.} {bandu} {barula} {mutsinzu} {tsiabo} {khutsia} {khwilukhanyinya} {bihubi.} {Bandu} {bene} {yabo} {bitsa} {nende} {marimu,} {tsindoboshi} {nende} {tsimpanga} {khukwana} {bihubi.} {Shibanyola} {bihubi} {tawe.} {Mbalaka} {yaranga} {khusakha} {na} {abola} {ndi} {yenyie} {khulola} {shia} {bandu} {bakhola.} {Inyanga} {yindi} {khandi} {yalanjilitsa} {bandu} {bahubanga} {tsing’ombe.} {Bandu} {khandi} {bitsa} {ni} {banyola} {Mbalaka} {anini} {mumusala} {asakhanga.} {Lwa} {khabaka} {yalanjilitsa} {bandu} {shibayitsa} {tawe} {shichila} {bapara} {ndinagwa} {ababehanga} {khuli} {matukhu} {kandi.} {Mbalaka} {yalanjilitsa} {mpaka} {nachioka.} {Mundu} {shiyayitsa} {tawe.} {Ikhabi} {indamanu,} {litikhu} {yili} {bihubi} {biali} {ni} {bitsi} {nibihuba} {tsingo’mbe} {tsia} {kuka} {wewe.}
{xa-le} {j-\'a:-liho} {mw-ana} {mu-jaji} {li-ra} {li-e} {mbalaka} {j-aj-il-a\engma{}a} {()-kuka} {w-e-w-e} {tsi\engma{}-ombe} {li-tuxu} {li-ndi} {\beta{}a-ndu} {ba-a-hulil-a} {na-a-land\ezh{}ilits-a\engma{}ga} {a-ri} {\beta{}a-ndu} {\beta{}a-i:ts-i} {xu-hu\beta{}-a} {tsi\engma{}-ombe} {\beta{}a-ndu} {\beta{}a-a-rul-a} {mu-tsin-dzu} {tsi-a-\beta{}o} {xu-tsi-a} {xw-iluxa-\palnas{}i\palnas{}-a} {\beta{}i-hu\beta{}i} {\beta{}a-ndu} {\beta{}a-ene} {j-a\beta{}-o} {\beta{}a-a-its-a} {nende} {ma-rimu} {tsin-doboshi} {nende} {tsim-pa\engma{}ga} {xu-kwan-a} {\beta{}i-hu\beta{}i} {\esh{}i-\beta{}a-\'a:-\palnas{}ol-a} {\beta{}i-hu\beta{}i} {t\'a:we} {mbalaka} {j-\'a:-ra\engma{}g-a} {xu-sax-a} {na} {a-\beta{}ol-a} {ndi} {j-e\palnas{}-i-\'e} {xu-lol-a} {\esh{}ja} {\beta{}a-ndu} {\beta{}\'a-xol-\'a} {i-\palnas{}a\engma{}ga} {j-indi} {xa:ndi} {j-\'a:-land\ezh{}ilits-a} {\beta{}a-ndu} {\beta{}a-hu\beta{}a\engma{}g-a} {tsi\engma{}-ombe} {\beta{}a-ndu} {xa:ndi} {\beta{}a-a-its-a} {ni} {\beta{}a-a-\palnas{}ol-a} {mbalaka} {a-nin-i} {mu-mu-sala} {a-sax-a\engma{}ga} {lwa} {xa-baka} {j-a:-land\ezh{}ilits-a} {\beta{}a-ndu} {\esh{}i-\beta{}a-a-jits-a} {t\'a:we} {\esh{}it\esh{}ila} {\beta{}a-a-par-a} {ndina\engma{}wa} {a-\beta{}a-\beta{}eh-a\engma{}ga} {xuli} {ma-tuxu} {xa:ndi} {mbalaka} {j-a-land\ezh{}ilits-a} {mbaka} {n\'{a}-\`a-t\esh{}iok-a.} {mu-ndu} {\esh{}i-j-a:-jits-a} {t\'a:we} {i-xa\beta{}i} {in-damanu,} {li-tixu} {j-il-i} {\beta{}i-hu\beta{}i} {\beta{}i-al-i} {ni} {\beta{}i-tsi-i} {ni-\beta{}i-hu\beta{}-a} {tsi\engma{}-ombe} {tsi-a} {(mu)-kuka} {w-e-w-e}
{12.long} {?.RPST.?} {1.child} {1.young} {5.name} {5.3rdPOS} {name} {3sg.graze.BEN.IMPF} {(1).grandfather} {1.3rdPOS} {10.cow} {5.day} {5.another} {2.person} {3pl.RPST.hear.FV} {CONN.3sg.shout.IMPF} {3sg.that?} {2.person} {3pl.come.NPST} {15.steal.FV} {10.cow} {2.person} {3pl.RPST.come-from.FV} {18.10.house} {10.V.3plPOS} {15.go.FV} {15.run.out.FV} {8.thief} {2.person} {2.same} {PDEM.2.V} {3pl.RPST.come.FV} {and} {6.spear} {10.whip} {and} {10.machete} {15.fight.FV} {8.thief} {NEG.3pl.RPST.find.FV} {8.thief} {NEG} {name} {3sg.RPST.start.FV} {15.laugh.FV} {and} {3sg.tell.FV} {?} {3sg.want.NPST.IMPF} {15.see.FV} {what} {2.person} {3pl.do.FV} {9.day} {9.indi} {again} {3sg-RPST-shout-FV} {2.person} {3sg-steal-FV} {10.cow} {2.person} {again} {3pl.RPST.come.FV} {and} {3pl.RPST.find.FV} {name} {3sg-climb-NPST} {18.3.tree} {3sg.laugh.FV} {\#counter?} {12?.three} {3sg.RPST.shout.FV} {2.person} {NEG.3pl.RPST.come.FV} {NEG} {because} {3pl.RPST.think.FV} {that?} {3sg.3plOBJ.lie.IMPF} {like?} {6.day} {other?} {name} {3sg-RPST-shout-FV} {until?} {CONN.3sg.tire.FV} {1.person} {NEG.3sg.RPST.come.FV} {NEG} {9.luck} {9.bad} {5.day} {PDEM.5.V} {8.thief} {8.be.NPST} {CONN} {8.come.NPST} {CONN.8.steal.FV} {10.cow} {9.?} {(1).grandfather} {1.3rdPOS}
\glt
\glend

Explanation of notation:
\begin{itemize}
  \setlength{\itemsep}{-1pt}
\item Number alone: showing noun class or noun class agreement
\item ? : currently unknown
\item 1st, 2nd, 3rd : first, second, or third person
\item sg, pl : singular, plural
\item POS : possessive
\item V : vowel
\item NEG : negative marker
\item OBJ : object
\item CONN : connector
\item RPST: Recent past tense
\item NPST : near past tense
\item BEN : benefactive
\item IMPF : imperfective
\item PDEM : proximal demonstrative
\item FV : final vowel
\end{itemize}